---
title: 'Part 2: chi-square with lizards'' tails'
author: "Katie Pietrucha"
date: "11/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(here)
library(janitor)
library(broom)
```

## Read in data
```{r}
lizards <- read_csv(here("data", "lizards.csv"))
```

Get counts of lizards by species
```{r}
lizards %>%
  count(spp) %>%
  arrange(-n) # order in decreasing order by count
```

## Make a subset

Keep:
- UTST lizards
- sites: IBPE (enclosure grassland), RABB (rabbit mesquite site), EAST (east tarbush site)
- tail condition is recorded (W = whole, B = broken)

```{r}
utst <- lizards %>%
  filter(spp == "UTST") %>%
  filter(site %in% c("IBPE", "RABB", "EAST")) %>%
  filter(tail %in% c("B", "W"))
```

## Find counts and proportions of lizards with broken or whole tails at each site.

```{r}
utst %>%
  count(site, tail) # but it's a long table
```

```{r}
tail_counts <- utst %>%
  janitor::tabyl(site, tail) # gets close to contingency table format

tail_proportions <- tail_counts %>%
  janitor::adorn_percentages() %>% # default is to calculate proportions by rows
  janitor::adorn_pct_formatting(digits = 2) %>%
  janitor::adorn_ns() # adds counts to percentages, great for communication
```
 
Is there a significant difference in these propprtions? Chi-square will answer. But nice to have counts and percentages reported together use janitor package stuff.

Do I have enough evidence to reject null hypothesis that tail breakage is independent of site?

## Convert to a contingency table

Use `colume_to_rowname()` to convert site column into row names so thats counts are the only data in the table.

```{r}
lizard_ct <- tail_counts %>% 
  column_to_rownames(var = "site")
```

## Chi-square

Are tail breakage and site truly independent? (would expect similar proportions)

```{r}
lizard_x2 <- chisq.test(lizard_ct)
lizard_x2

lizard_tidy <- broom::tidy(lizard_x2) # tidy data fram to call components from for in line referencing.
```

Degrees of freedom for chi-square = (number of rows - 1) * (number of columns - 1)

Yes, we have sufficient evidence to reject null hypothesis of independence. Retain alt hypothesis that  site and tail breakage are not independent.There is a significant association between site and tail breakage. 

Using inline referencing:

There is a significant association (i.e. non-independence) between site and lizard tail breakage ($\chi$^2^(`r lizard_tidy$parameter`)) = `r lizard_tidy$statistic`, *p* = `r lizard_tidy$p.value`).
